\chapter*{Acknowledgements}
\markboth{Acknowledgements}{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
% put your text here
The completion of this thesis would not have been possible without the unwavering support and guidance of numerous individuals who have played a significant role in my academic journey.

\vspace{2ex}
First and foremost, I extend my deepest gratitude to my supervisor, Prof. François Fleuret, for their unwavering mentorship and belief in my abilities. Their insightful guidance, constructive feedback, and relentless encouragement have been instrumental in shaping my research and ensuring its clarity and rigor. I am deeply indebted to their expertise and dedication.

\vspace{2ex}
Secondly, I am deeply appreciative of the insightful comments, intellectually stimulating discussions, and generous sharing of knowledge and expertise provided by the researchers and coauthors at ams OSRAM, Hugues Salamin, Camilla Carta, and Yann Lepoittevin. Their insights and collaborative spirit were invaluable in enriching my research experience and expanding my understanding of the subject matter.

\vspace{2ex}
In addition, I would like to express my sincere gratitude to the entire Idiap team, in particular, my co-supervisor Prof. Daniel Gatica-Perez, for creating a supportive and enriching environment for my PhD studies.

\vspace{2ex}
I am also thankful to Jan Rüegg and Sebastian Martin from Apple Zurich, whose guidance and collaboration during my internship profoundly enriched my proficiency in 3D rendering and familiarized me with the intricacies of the industry environment.

\vspace{2ex}
Last but not least, I extend my heartfelt thanks to my family and friends for their unwavering love, support, and encouragement throughout this challenging yet rewarding journey. Their belief in me served as a constant source of motivation, and their unwavering support helped me persevere through moments of doubt and self-doubt.

\bigskip
 
\noindent\textit{Lausanne, \today}
\hfill S.~Johari
